---
title: "Machine Learning Assignment"
author: "François de BELLEFON"
date: "2017, October 15th"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE);
library(caret)
library(AppliedPredictiveModeling)
library(plyr)
library(dplyr)
```
#Overview
We are going to build a model to predict the manner in which a weight lifting exercice is done, from data collected from sports devices which feature accelerometers.

#Reproducibility : setting seed
Let's set a seed for reproducibility purposes :
```{r seedSetting}
set.seed(42)
```


#Getting the data
Let's load the data that we will use to train the model.
```{r dataDownload}
exerciseData <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv")
```





#Taking a look at our data
First of all, let's take a look at our dara :
```{r dataSummary}
str(exerciseData)
```

##Columns classifications
We can categorise the data in 3 categories :

- Data identifying the observation (Data regarding the test index, the user name, the timestamp and the fact that we are or not in a new time window. )
- Raw measurements
- Data derived from the raw mesurements (total, kurtosis, skewness, minimal, maximal, amplitude, variance, standard deviation, average)


##Missing data analysis
As we can see, there are a lot of missing data (NA or empty).

Let's see which columns have more than 90% missing data :
```{r isDataMissing}
missingDataCols <- names(exerciseData[which(colSums(exerciseData == '' | is.na(exerciseData))/nrow(exerciseData)*100 > 90)])
missingDataCols
```
As we can see, the columns with the most missing data are not the raw data, byt the ones which are derived from the raw data.


##Class imbalance analysis

Let's also see how balanced our data classes are :
```{r isDataBalanced}
table(exerciseData$classe)
```
As we can see, class "A" is over represented in our sample. It may lead to imbalance when training our model. We will deal with it when training it.



#Model building and testing
Now, let's get to the proper building and testing of our model.


##Training and testing set
Let's divide this data into a training set (which we will use to train our model) and a test set (which we will use to see if our model is correct).
```{r dataPartition}
inTrain = createDataPartition(exerciseData$classe, p = 3/4)[[1]]
tr <- exerciseData[inTrain,]
te <- exerciseData[-inTrain,]
```

## Model Building
### Problem type and model selection
We have a supervised classification problem : we want to categorize new data into determined categories.

We are going to use Gradient Boosting Model (gbm).

###Feature selection

Let's now select the feature that we will use.

As we saw before when examining our data :

- Some columns are only there to identify the measurement and are not relevant as features.
- Some columns are derived from the raw data ; so keeping them in addition to the raw data may lead to unnecessary variance. Also, we saw when examining our data that the derived data had a majority of NA and empty values ; so we remove these derived data.

Let's actually remove these columns :


```{r trainingFeaturesSelection}
tr2 <- select(tr,-starts_with("var"),-starts_with("min"),-starts_with("max"),-starts_with("avg"),-starts_with("skewness"),-starts_with("kurtosis"),-starts_with("stddev"),-starts_with("amplitude"),-matches("X"),-(1:7))
```

###Data preprocessing
####Cross Validation
We are going to chose a way to validate our model.

We'll use k-fold Cross Validation : we will devide the training data in 10 equal sized subsamples, train the model with 9 of these 10 subsamples, validate in on the 10th, and do it again with the other subsamples.

####Class imbalance
As we have seen before, class A is more represented than the others.

We are going to deal with this class imbalance by  using upsampling in the control training parameters : this randomly replicates instances of the minority classes.


Let's implement this data preprocessing :

```{r controlParameter}
Ctrl <- trainControl(method="cv",number=10,sampling="up")
```

###Training the model

Now, let's actually train our gbm model with our selected features, our oversampling options and the k-fold validation as mentionned before :
```{r modelTraining}
modFit <- train(classe~.,method="gbm",trControl=Ctrl,data=tr2,verbose=FALSE)
```

#Model evaluation
##Evaluation on the test set
Let's do a prediction with the test set (te) that we defined earlier, and evaluate the result in comparison with the actual values :
```{r modelEvaluation}
predicted <- predict(modFit,newdata = te)
confusionMatrix(predicted,te$class)
```


As we can see, we have an accuracy of 96% ; for each class, we have more that 90% sensitivity and specificity.

#Prediction for 20 cases
Now, let's use our model to predict the 20 cases as requested :
```{r 20CasesEvaluation}
#Test set
testData <- read.csv("https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv")
predict(modFit,testData)
```